---
title: "RNA-seq DESeq2 Differential Expression and QC Workflow"
author: "Liz Dauster"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 7
)

set.seed(42)
```

# Overview

This workflow performs RNA-seq quality control, documented outlier handling, differential expression analysis, and result export for Gene Expression Omnibus accession GSE203539.

The analysis starts from the combined count matrix in `data/all_counts.txt`, performs quality control on all samples, explicitly flags pre-specified outlier samples, compares all-sample and filtered DESeq2 results through sensitivity analysis, and then proceeds with the filtered DESeq2 workflow.

This analysis is intended as a reproducible reporting workflow. It does not claim biological conclusions beyond the differential expression outputs generated here.

# Analysis plan

1. Load the count matrix from `data/all_counts.txt`.
2. Parse sample metadata from sample names.
3. Perform QC on all samples before filtering.
4. Explicitly flag the following outlier samples:
   - `AEC.mock.III`
   - `AEC.WSN.PP1.I`
   - `AM.mock.II`
   - `AM.WSN.III`
5. Run a sensitivity analysis comparing all-sample and filtered DESeq2 results.
6. Continue with filtered DESeq2 analysis.
7. Export QC figures, DEG result tables, and summary tables to `results/`.

# Load packages

```{r load-packages}
required_packages <- c(
  "DESeq2",
  "EnhancedVolcano",
  "org.Mm.eg.db",
  "AnnotationDbi",
  "ggplot2",
  "pheatmap",
  "dplyr",
  "tidyr",
  "readr",
  "stringr",
  "tibble",
  "purrr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required package(s): ",
    paste(missing_packages, collapse = ", "),
    "\nInstall missing packages before knitting this workflow."
  )
}

library(DESeq2)
library(EnhancedVolcano)
library(org.Mm.eg.db)
library(AnnotationDbi)
library(ggplot2)
library(pheatmap)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(tibble)
library(purrr)
```

# Project paths

```{r paths}
data_dir <- "data"
results_dir <- "results"
figures_dir <- file.path(results_dir, "figures")
tables_dir <- file.path(results_dir, "tables")
deg_dir <- file.path(results_dir, "deg_outputs")
qc_dir <- file.path(results_dir, "qc")

walk(
  c(results_dir, figures_dir, tables_dir, deg_dir, qc_dir),
  ~ dir.create(.x, recursive = TRUE, showWarnings = FALSE)
)

counts_file <- file.path(data_dir, "all_counts.txt")
annotation_file <- file.path(data_dir, "GRCm39_annotation.txt")

if (!file.exists(counts_file)) {
  stop("Missing required count matrix: ", counts_file)
}
```

# Import count matrix

```{r import-counts}
raw_counts <- read.delim(
  counts_file,
  header = TRUE,
  sep = "\t",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

if (ncol(raw_counts) < 3) {
  stop("Count matrix appears too small. Expected one gene ID column plus sample count columns.")
}

# Support both simple count matrices and featureCounts-style tables.
featurecounts_annotation_cols <- c("Chr", "Start", "End", "Strand", "Length")
possible_gene_cols <- c("gene_id", "Geneid", "GeneID", "Gene.stable.ID", "gene", "id")

gene_col <- intersect(possible_gene_cols, names(raw_counts))[1]

if (is.na(gene_col)) {
  gene_col <- names(raw_counts)[1]
  message("No standard gene ID column found. Using first column as gene_id: ", gene_col)
}

counts_tbl <- raw_counts %>%
  rename(gene_id = all_of(gene_col)) %>%
  select(-any_of(featurecounts_annotation_cols))

# If featureCounts used full BAM paths as sample names, simplify them where possible.
clean_sample_name <- function(x) {
  x %>%
    basename() %>%
    str_remove("\\.bam$") %>%
    str_remove("\\.sam$") %>%
    str_remove("\\.sorted$")
}

names(counts_tbl) <- c("gene_id", clean_sample_name(names(counts_tbl)[-1]))

sample_cols <- setdiff(names(counts_tbl), "gene_id")

counts_tbl <- counts_tbl %>%
  mutate(across(all_of(sample_cols), as.numeric))

if (any(is.na(counts_tbl[sample_cols]))) {
  stop("Count matrix contains NA values after numeric conversion. Check input formatting.")
}

count_matrix <- counts_tbl %>%
  group_by(gene_id) %>%
  summarise(across(all_of(sample_cols), ~ sum(.x, na.rm = TRUE)), .groups = "drop") %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

storage.mode(count_matrix) <- "integer"

if (any(count_matrix < 0)) {
  stop("Count matrix contains negative values, which are invalid for DESeq2.")
}

cat("Genes:", nrow(count_matrix), "\n")
cat("Samples:", ncol(count_matrix), "\n")
```

# Build sample metadata

```{r sample-metadata}
outlier_samples <- c(
  "AEC.mock.III",
  "AEC.WSN.PP1.I",
  "AM.mock.II",
  "AM.WSN.III"
)

metadata <- tibble(sample_id = colnames(count_matrix)) %>%
  mutate(
    cell_type = case_when(
      str_starts(sample_id, "AEC\\.") ~ "AEC",
      str_starts(sample_id, "AM\\.") ~ "AM",
      TRUE ~ NA_character_
    ),
    condition = case_when(
      str_detect(sample_id, "\\.mock\\.") ~ "mock",
      str_detect(sample_id, "\\.WSN\\.PP1\\.") ~ "WSN_PP1",
      str_detect(sample_id, "\\.WSN\\.PP\\.") ~ "WSN_PP1",
      str_detect(sample_id, "\\.WSN\\.") ~ "WSN",
      TRUE ~ NA_character_
    ),
    replicate = str_extract(sample_id, "[^.]+$"),
    outlier_flag = sample_id %in% outlier_samples
  ) %>%
  mutate(
    cell_type = factor(cell_type, levels = c("AEC", "AM")),
    condition = factor(condition, levels = c("mock", "WSN", "WSN_PP1")),
    replicate = factor(replicate),
    group = factor(paste(cell_type, condition, sep = "."))
  )

if (any(is.na(metadata$cell_type)) || any(is.na(metadata$condition))) {
  print(metadata)
  stop("Unable to infer cell_type or condition for one or more samples. Check sample naming conventions.")
}

missing_outliers <- setdiff(outlier_samples, metadata$sample_id)

if (length(missing_outliers) > 0) {
  warning("Pre-specified outlier sample(s) not found in count matrix: ", paste(missing_outliers, collapse = ", "))
}

if (!all(metadata$sample_id == colnames(count_matrix))) {
  stop("Metadata sample order does not match count matrix column order.")
}

write_csv(metadata, file.path(qc_dir, "sample_metadata_with_outlier_flags.csv"))
metadata
```

# Optional gene annotation

```{r read-annotation}
if (file.exists(annotation_file)) {
  annotation <- read.delim(
    annotation_file,
    header = TRUE,
    sep = "\t",
    quote = "\"",
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  if ("Gene.stable.ID" %in% names(annotation)) {
    annotation_gene <- annotation[!duplicated(annotation$Gene.stable.ID), ]
  } else {
    warning("Annotation file found, but Gene.stable.ID column was not detected. Results will be exported without annotation merge.")
    annotation_gene <- NULL
  }
} else {
  warning("Optional annotation file not found: ", annotation_file, ". Results will be exported without annotation merge.")
  annotation_gene <- NULL
}
```

# QC on all samples

Quality control is performed before excluding any samples. The outlier samples are flagged visually and retained in this section.

## Library size

```{r qc-library-size-all}
library_sizes <- metadata %>%
  mutate(total_counts = colSums(count_matrix)[sample_id])

p_library <- ggplot(library_sizes, aes(x = reorder(sample_id, total_counts), y = total_counts, fill = outlier_flag)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Library size by sample: all samples",
    x = "Sample",
    y = "Total counts",
    fill = "Flagged outlier"
  ) +
  theme_bw(base_size = 12)

p_library

ggsave(file.path(figures_dir, "qc_library_size_all_samples.png"), p_library, width = 9, height = 6, dpi = 300)
write_csv(library_sizes, file.path(qc_dir, "library_sizes_all_samples.csv"))
```

## Variance-stabilized counts

```{r qc-vst-all}
keep_genes_qc <- rowSums(count_matrix >= 10) >= 2

if (sum(keep_genes_qc) < 100) {
  stop("Too few genes pass the QC expression filter. Check count matrix input.")
}

dds_qc_all <- DESeqDataSetFromMatrix(
  countData = count_matrix[keep_genes_qc, ],
  colData = metadata %>% as.data.frame() %>% column_to_rownames("sample_id"),
  design = ~ cell_type + condition
)

vsd_qc_all <- vst(dds_qc_all, blind = TRUE)
```

## PCA: all samples

```{r qc-pca-all}
pca_all <- plotPCA(vsd_qc_all, intgroup = c("cell_type", "condition"), returnData = TRUE) %>%
  rownames_to_column("sample_id") %>%
  left_join(metadata, by = "sample_id")

percent_var_all <- round(100 * attr(pca_all, "percentVar"))

p_pca_all <- ggplot(pca_all, aes(x = PC1, y = PC2, color = condition, shape = cell_type, label = sample_id)) +
  geom_point(size = 3) +
  geom_text(aes(alpha = outlier_flag), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_alpha_manual(values = c(`FALSE` = 0.45, `TRUE` = 1)) +
  labs(
    title = "PCA of variance-stabilized counts: all samples",
    x = paste0("PC1: ", percent_var_all[1], "% variance"),
    y = paste0("PC2: ", percent_var_all[2], "% variance")
  ) +
  theme_bw(base_size = 12) +
  coord_fixed()

p_pca_all

ggsave(file.path(figures_dir, "qc_pca_all_samples.png"), p_pca_all, width = 9, height = 7, dpi = 300)
write_csv(pca_all, file.path(qc_dir, "pca_all_samples.csv"))
```

## Sample distance heatmap: all samples

```{r qc-distance-all}
sample_dist_all <- dist(t(assay(vsd_qc_all)))
sample_dist_mat_all <- as.matrix(sample_dist_all)
rownames(sample_dist_mat_all) <- colnames(count_matrix)
colnames(sample_dist_mat_all) <- colnames(count_matrix)

annotation_col_all <- metadata %>%
  select(sample_id, cell_type, condition, outlier_flag) %>%
  as.data.frame() %>%
  column_to_rownames("sample_id")

png(file.path(figures_dir, "qc_sample_distance_all_samples.png"), width = 2200, height = 1800, res = 300)
pheatmap(
  sample_dist_mat_all,
  clustering_distance_rows = sample_dist_all,
  clustering_distance_cols = sample_dist_all,
  annotation_col = annotation_col_all,
  main = "Sample distance heatmap: all samples"
)
dev.off()

pheatmap(
  sample_dist_mat_all,
  clustering_distance_rows = sample_dist_all,
  clustering_distance_cols = sample_dist_all,
  annotation_col = annotation_col_all,
  main = "Sample distance heatmap: all samples"
)
```

# Helper functions

```{r helper-functions}
make_dds_for_cell <- function(counts, sample_metadata, cell, remove_outliers = FALSE) {
  meta <- sample_metadata %>%
    filter(cell_type == cell)

  if (remove_outliers) {
    meta <- meta %>% filter(!outlier_flag)
  }

  mat <- counts[, meta$sample_id, drop = FALSE]
  keep <- rowSums(mat >= 10) >= 2
  mat <- mat[keep, , drop = FALSE]

  meta <- meta %>%
    mutate(condition = droplevels(condition))

  if (nlevels(meta$condition) < 2) {
    stop("Fewer than two treatment conditions available for ", cell, ".")
  }

  dds <- DESeqDataSetFromMatrix(
    countData = mat,
    colData = meta %>% as.data.frame() %>% column_to_rownames("sample_id"),
    design = ~ condition
  )

  DESeq(dds)
}

extract_contrast <- function(dds, cell, contrast_label, numerator, denominator, alpha = 0.05) {
  available_conditions <- unique(as.character(colData(dds)$condition))

  if (!all(c(numerator, denominator) %in% available_conditions)) {
    warning("Skipping ", contrast_label, ": missing required condition.")
    return(NULL)
  }

  res <- results(
    dds,
    contrast = c("condition", numerator, denominator),
    alpha = alpha
  )

  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    mutate(
      cell_type = cell,
      contrast = contrast_label,
      numerator = numerator,
      denominator = denominator,
      strict_deg = !is.na(padj) & padj <= 0.005 & abs(log2FoldChange) > 2,
      regulation = case_when(
        strict_deg & log2FoldChange > 0 ~ "up",
        strict_deg & log2FoldChange < 0 ~ "down",
        TRUE ~ "not_strict"
      )
    ) %>%
    relocate(cell_type, contrast, gene_id)
}

run_cell_analysis <- function(counts, sample_metadata, cell, analysis_set, remove_outliers = FALSE) {
  dds <- make_dds_for_cell(counts, sample_metadata, cell, remove_outliers = remove_outliers)

  comparisons <- tibble(
    contrast = c(
      paste0(cell, "_WSN_vs_mock"),
      paste0(cell, "_WSN_PP1_vs_WSN")
    ),
    numerator = c("WSN", "WSN_PP1"),
    denominator = c("mock", "WSN")
  )

  results_tbl <- pmap_dfr(
    comparisons,
    ~ extract_contrast(dds, cell = cell, contrast_label = ..1, numerator = ..2, denominator = ..3)
  ) %>%
    mutate(analysis_set = analysis_set) %>%
    relocate(analysis_set, .after = contrast)

  list(dds = dds, results = results_tbl)
}

summarize_degs <- function(results_tbl) {
  results_tbl %>%
    group_by(analysis_set, cell_type, contrast) %>%
    summarise(
      genes_tested = n(),
      padj_0_05 = sum(!is.na(padj) & padj <= 0.05),
      padj_0_005 = sum(!is.na(padj) & padj <= 0.005),
      strict_degs = sum(strict_deg, na.rm = TRUE),
      strict_up = sum(strict_deg & log2FoldChange > 0, na.rm = TRUE),
      strict_down = sum(strict_deg & log2FoldChange < 0, na.rm = TRUE),
      .groups = "drop"
    )
}

annotate_results <- function(results_tbl, annotation_gene) {
  if (is.null(annotation_gene)) {
    return(results_tbl)
  }

  results_tbl %>%
    left_join(annotation_gene, by = c("gene_id" = "Gene.stable.ID"))
}
```

# Sensitivity analysis: all-sample vs filtered DESeq2 results

This section evaluates whether the pre-specified outlier handling decision changes differential expression outputs. Both all-sample and filtered results are exported for transparency.

```{r sensitivity-run}
all_sample_analyses <- map(
  c("AEC", "AM"),
  ~ run_cell_analysis(count_matrix, metadata, cell = .x, analysis_set = "all_samples", remove_outliers = FALSE)
)

filtered_analyses <- map(
  c("AEC", "AM"),
  ~ run_cell_analysis(count_matrix, metadata, cell = .x, analysis_set = "filtered", remove_outliers = TRUE)
)

all_sample_results <- map_dfr(all_sample_analyses, "results")
filtered_results <- map_dfr(filtered_analyses, "results")

sensitivity_results <- bind_rows(all_sample_results, filtered_results)
sensitivity_summary <- summarize_degs(sensitivity_results)

write_csv(sensitivity_results, file.path(deg_dir, "deseq2_results_all_vs_filtered.csv"))
write_csv(sensitivity_summary, file.path(tables_dir, "deg_summary_all_vs_filtered.csv"))

sensitivity_summary
```

```{r sensitivity-gene-level-comparison}
sensitivity_gene_comparison <- sensitivity_results %>%
  select(cell_type, contrast, analysis_set, gene_id, log2FoldChange, padj, strict_deg) %>%
  pivot_wider(
    names_from = analysis_set,
    values_from = c(log2FoldChange, padj, strict_deg),
    names_sep = "__"
  ) %>%
  mutate(
    strict_deg__all_samples = replace_na(strict_deg__all_samples, FALSE),
    strict_deg__filtered = replace_na(strict_deg__filtered, FALSE),
    delta_log2fc = log2FoldChange__filtered - log2FoldChange__all_samples,
    strict_status = case_when(
      strict_deg__all_samples & strict_deg__filtered ~ "strict_in_both",
      strict_deg__all_samples & !strict_deg__filtered ~ "strict_all_only",
      !strict_deg__all_samples & strict_deg__filtered ~ "strict_filtered_only",
      TRUE ~ "not_strict_in_either"
    )
  )

sensitivity_overlap_summary <- sensitivity_gene_comparison %>%
  group_by(cell_type, contrast, strict_status) %>%
  summarise(n_genes = n(), .groups = "drop")

write_csv(sensitivity_gene_comparison, file.path(tables_dir, "sensitivity_gene_level_comparison.csv"))
write_csv(sensitivity_overlap_summary, file.path(tables_dir, "sensitivity_overlap_summary.csv"))

sensitivity_overlap_summary
```

```{r sensitivity-plot}
p_sensitivity <- sensitivity_gene_comparison %>%
  filter(!is.na(log2FoldChange__all_samples), !is.na(log2FoldChange__filtered)) %>%
  ggplot(aes(x = log2FoldChange__all_samples, y = log2FoldChange__filtered, color = strict_status)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ contrast, scales = "free") +
  labs(
    title = "Sensitivity analysis: all-sample vs filtered log2 fold change",
    x = "All-sample log2 fold change",
    y = "Filtered log2 fold change",
    color = "Strict DEG status"
  ) +
  theme_bw(base_size = 12)

p_sensitivity

ggsave(file.path(figures_dir, "sensitivity_log2fc_all_vs_filtered.png"), p_sensitivity, width = 10, height = 7, dpi = 300)
```

# Filtered DESeq2 workflow

The remaining workflow uses the filtered dataset, excluding the explicitly flagged outlier samples.

```{r filtered-metadata}
metadata_filtered <- metadata %>%
  filter(!outlier_flag)

count_matrix_filtered <- count_matrix[, metadata_filtered$sample_id, drop = FALSE]

write_csv(metadata_filtered, file.path(qc_dir, "sample_metadata_filtered.csv"))
metadata_filtered
```

## Filtered PCA

```{r filtered-vst-pca}
keep_genes_filtered <- rowSums(count_matrix_filtered >= 10) >= 2

dds_qc_filtered <- DESeqDataSetFromMatrix(
  countData = count_matrix_filtered[keep_genes_filtered, ],
  colData = metadata_filtered %>% as.data.frame() %>% column_to_rownames("sample_id"),
  design = ~ cell_type + condition
)

vsd_qc_filtered <- vst(dds_qc_filtered, blind = TRUE)

pca_filtered <- plotPCA(vsd_qc_filtered, intgroup = c("cell_type", "condition"), returnData = TRUE) %>%
  rownames_to_column("sample_id") %>%
  left_join(metadata_filtered, by = "sample_id")

percent_var_filtered <- round(100 * attr(pca_filtered, "percentVar"))

p_pca_filtered <- ggplot(pca_filtered, aes(x = PC1, y = PC2, color = condition, shape = cell_type, label = sample_id)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, alpha = 0.7) +
  labs(
    title = "PCA of variance-stabilized counts: filtered samples",
    x = paste0("PC1: ", percent_var_filtered[1], "% variance"),
    y = paste0("PC2: ", percent_var_filtered[2], "% variance")
  ) +
  theme_bw(base_size = 12) +
  coord_fixed()

p_pca_filtered

ggsave(file.path(figures_dir, "qc_pca_filtered_samples.png"), p_pca_filtered, width = 9, height = 7, dpi = 300)
write_csv(pca_filtered, file.path(qc_dir, "pca_filtered_samples.csv"))
```

## Filtered sample distance heatmap

```{r filtered-distance-heatmap}
sample_dist_filtered <- dist(t(assay(vsd_qc_filtered)))
sample_dist_mat_filtered <- as.matrix(sample_dist_filtered)
rownames(sample_dist_mat_filtered) <- colnames(count_matrix_filtered)
colnames(sample_dist_mat_filtered) <- colnames(count_matrix_filtered)

annotation_col_filtered <- metadata_filtered %>%
  select(sample_id, cell_type, condition) %>%
  as.data.frame() %>%
  column_to_rownames("sample_id")

png(file.path(figures_dir, "qc_sample_distance_filtered_samples.png"), width = 2200, height = 1800, res = 300)
pheatmap(
  sample_dist_mat_filtered,
  clustering_distance_rows = sample_dist_filtered,
  clustering_distance_cols = sample_dist_filtered,
  annotation_col = annotation_col_filtered,
  main = "Sample distance heatmap: filtered samples"
)
dev.off()

pheatmap(
  sample_dist_mat_filtered,
  clustering_distance_rows = sample_dist_filtered,
  clustering_distance_cols = sample_dist_filtered,
  annotation_col = annotation_col_filtered,
  main = "Sample distance heatmap: filtered samples"
)
```

# Export filtered DEG results

```{r filtered-result-exports}
filtered_results_annotated <- annotate_results(filtered_results, annotation_gene)
filtered_summary <- summarize_degs(filtered_results)
strict_filtered_results <- filtered_results_annotated %>%
  filter(strict_deg) %>%
  arrange(cell_type, contrast, padj, desc(abs(log2FoldChange)))

write_csv(filtered_results_annotated, file.path(deg_dir, "deseq2_results_filtered_annotated.csv"))
write_csv(strict_filtered_results, file.path(deg_dir, "strict_deg_results_filtered_padj005_log2fc2.csv"))
write_csv(filtered_summary, file.path(tables_dir, "deg_summary_filtered.csv"))

# Export one annotated table per contrast for downstream Python reporting.
filtered_results_annotated %>%
  group_split(contrast) %>%
  walk(function(tbl) {
    contrast_name <- unique(tbl$contrast)
    output_file <- file.path(deg_dir, paste0(contrast_name, "_DESeq2_annot.csv"))
    write_csv(tbl, output_file)
  })

filtered_summary
```

# Filtered DEG visualizations

## DEG count summary

```{r deg-count-plot}
p_deg_counts <- filtered_summary %>%
  select(cell_type, contrast, strict_up, strict_down) %>%
  pivot_longer(cols = c(strict_up, strict_down), names_to = "direction", values_to = "n_genes") %>%
  ggplot(aes(x = contrast, y = n_genes, fill = direction)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Strict DEG counts in filtered analysis",
    x = "Contrast",
    y = "Number of genes",
    fill = "Direction"
  ) +
  theme_bw(base_size = 12)

p_deg_counts

ggsave(file.path(figures_dir, "strict_deg_counts_filtered.png"), p_deg_counts, width = 8, height = 5, dpi = 300)
```

## Volcano plots

```{r volcano-plots}
make_volcano <- function(results_tbl, contrast_id) {
  plot_tbl <- results_tbl %>%
    filter(contrast == contrast_id) %>%
    mutate(
      padj_plot = if_else(is.na(padj), NA_real_, pmax(padj, 1e-300)),
      threshold_group = case_when(
        strict_deg & log2FoldChange > 0 ~ "strict_up",
        strict_deg & log2FoldChange < 0 ~ "strict_down",
        TRUE ~ "not_strict"
      )
    )

  ggplot(plot_tbl, aes(x = log2FoldChange, y = -log10(padj_plot), color = threshold_group)) +
    geom_point(alpha = 0.45, size = 1) +
    geom_vline(xintercept = c(-2, 2), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.005), linetype = "dashed") +
    labs(
      title = paste("Volcano plot:", contrast_id),
      x = "log2 fold change",
      y = "-log10 adjusted p-value",
      color = "Threshold group"
    ) +
    theme_bw(base_size = 12)
}

for (contrast_id in unique(filtered_results$contrast)) {
  p <- make_volcano(filtered_results, contrast_id)
  print(p)
  ggsave(file.path(figures_dir, paste0("volcano_", contrast_id, ".png")), p, width = 8, height = 6, dpi = 300)
}
```

## EnhancedVolcano plots with gene labels

```{r enhanced-volcano-plots}
make_enhanced_volcano <- function(dds, results_tbl, contrast_id, numerator, denominator) {
  res <- results(
    dds,
    contrast = c("condition", numerator, denominator)
  )

  res_shrunk <- lfcShrink(
    dds,
    contrast = c("condition", numerator, denominator),
    res = res,
    type = "normal"
  )

  symbols <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys = rownames(res_shrunk),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  labels <- symbols[match(rownames(res_shrunk), names(symbols))]
  labels[is.na(labels)] <- rownames(res_shrunk)[is.na(labels)]

  top_genes <- results_tbl %>%
    filter(contrast == contrast_id) %>%
    arrange(padj) %>%
    slice_head(n = 25) %>%
    pull(gene_id)

  EnhancedVolcano(
    res_shrunk,
    lab = labels,
    selectLab = labels[match(top_genes, rownames(res_shrunk))],
    x = "log2FoldChange",
    y = "padj",
    title = contrast_id,
    subtitle = paste(numerator, "vs", denominator),
    pCutoff = 0.005,
    FCcutoff = 2,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    pointSize = 2.0,
    labSize = 3.0
  )
}

filtered_dds_AEC <- filtered_analyses[[1]]$dds
filtered_dds_AM <- filtered_analyses[[2]]$dds

volcano_specs <- tibble(
  cell_type = c("AEC", "AEC", "AM", "AM"),
  contrast = c("AEC_WSN_vs_mock", "AEC_WSN_PP1_vs_WSN", "AM_WSN_vs_mock", "AM_WSN_PP1_vs_WSN"),
  numerator = c("WSN", "WSN_PP1", "WSN", "WSN_PP1"),
  denominator = c("mock", "WSN", "mock", "WSN")
)

for (i in seq_len(nrow(volcano_specs))) {
  spec <- volcano_specs[i, ]
  dds <- if (spec$cell_type == "AEC") filtered_dds_AEC else filtered_dds_AM

  png(file.path(figures_dir, paste0("enhanced_volcano_", spec$contrast, ".png")), width = 2200, height = 1800, res = 300)
  print(make_enhanced_volcano(dds, filtered_results, spec$contrast, spec$numerator, spec$denominator))
  dev.off()

  print(make_enhanced_volcano(dds, filtered_results, spec$contrast, spec$numerator, spec$denominator))
}
```

# Heatmaps of top DEGs

```{r heatmaps}
row_z_score <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(rep(0, length(x)))
  }
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

make_top_gene_heatmap <- function(vsd, results_tbl, contrast_id, metadata_subset, top_n = 25) {
  top_genes <- results_tbl %>%
    filter(contrast == contrast_id) %>%
    arrange(padj) %>%
    slice_head(n = top_n) %>%
    pull(gene_id)

  top_genes <- intersect(top_genes, rownames(assay(vsd)))

  if (length(top_genes) < 2) {
    warning("Too few genes available for heatmap: ", contrast_id)
    return(NULL)
  }

  heatmap_mat <- assay(vsd)[top_genes, , drop = FALSE]
  heatmap_mat_z <- t(apply(heatmap_mat, 1, row_z_score))

  symbols <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys = rownames(heatmap_mat_z),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  row_labels <- symbols[match(rownames(heatmap_mat_z), names(symbols))]
  row_labels[is.na(row_labels)] <- rownames(heatmap_mat_z)[is.na(row_labels)]
  rownames(heatmap_mat_z) <- make.unique(row_labels)

  annotation_col <- metadata_subset %>%
    select(sample_id, condition) %>%
    as.data.frame() %>%
    column_to_rownames("sample_id")

  pheatmap(
    heatmap_mat_z,
    annotation_col = annotation_col,
    main = paste("Top", top_n, "genes:", contrast_id),
    fontsize_row = 7
  )
}

for (cell in c("AEC", "AM")) {
  dds <- if (cell == "AEC") filtered_dds_AEC else filtered_dds_AM
  vsd <- vst(dds, blind = FALSE)
  meta_cell <- metadata_filtered %>% filter(cell_type == cell)

  for (contrast_id in unique(filtered_results$contrast[filtered_results$cell_type == cell])) {
    png(file.path(figures_dir, paste0("heatmap_top25_", contrast_id, ".png")), width = 2200, height = 1800, res = 300)
    make_top_gene_heatmap(vsd, filtered_results, contrast_id, meta_cell, top_n = 25)
    dev.off()

    make_top_gene_heatmap(vsd, filtered_results, contrast_id, meta_cell, top_n = 25)
  }
}
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
