---
title: "DESeq2 Differential Expression and QC Analysis"
author: "Liz Dauster"
date: "April 5, 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Adjust working directory as needed
knitr::opts_knit$set(root.dir = "C:/Users/lizda/Desktop/New School/BMB502/Term Project/selected counts")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 7)
```
# Differential Expression Analysis Influenca A Infection and 1PP1 Inhibitor Effect
in Mouse Alveolar Epithelial Cells (AECs) and Alveolar Macrophages (AMs)

This R Markdown file contains the final DESeq2 workflow for the influenza WSN infection
and 1NMPP1 inhibitor treatment analysis, Gene Expression Omnibus accession GSE203539.
The final analysis uses the cleaned dataset after sensitivity analysis and removal
of QC outlier samples.

Outlier samples removed prior to final downstream analysis:

- `AEC.mock.III`
- `AEC.WSN.PP1.I`
- `AM.mock.II`
- `AM.WSN.III`

A separate sensitivity analysis was performed with all original samples.
Because the outliers affected differential expression results, especially for AM contrasts,
the cleaned dataset was used for the final analysis.

# Load libraries

```{r load-libraries}
#*******************************************
# Load libraries
#*******************************************
library(DESeq2)
library(EnhancedVolcano)
library(org.Mm.eg.db)
library(ggplot2)
library(ComplexHeatmap)
```

# Read gene annotation table

```{r read-annotation}
#*******************************************
# Read in gene annotation table
#*******************************************
annotation <- read.table("GRCm39_annotation.txt", header = TRUE, sep = "\t", quote = "\"")

# Keep one annotation row per Ensembl gene ID for downstream merges
annotation_gene <- annotation[!duplicated(annotation$Gene.stable.ID), ]
```

# Main analysis: cleaned dataset without outliers

## Read count matrix

```{r read-counts}
#*******************************************
# Main Analysis: Original Samples without Outliers
#*******************************************

# Read in the raw read counts after excluding outlier samples
rawCounts <- read.table("selected_counts.txt", header = TRUE, sep = "\t")
head(rawCounts)

# Reformat so rownames are Ensembl Gene IDs
rnaseqMatrix <- rawCounts[, 2:15]
rownames(rnaseqMatrix) <- rawCounts[, 1]
head(rnaseqMatrix)
```

## Define sample metadata

```{r sample-metadata}
# Define sample metadata
samples <- data.frame(
  CellType = c(rep("AEC", 7), rep("AM", 7)),
  
  Treatment = c(rep("mock", 2), rep("WSN", 3), rep("WSN_PP1", 2),
                rep("mock", 2), rep("WSN", 2), rep("WSN_PP1", 3)),
  
  Replicate = c("I", "II", "I", "II", "III", "II", "III",
                "I", "III", "I", "II", "I", "II", "III"),
  
  row.names = c("AEC.mock.I", "AEC.mock.II",
                "AEC.WSN.I", "AEC.WSN.II", "AEC.WSN.III",
                "AEC.WSN.PP1.II", "AEC.WSN.PP1.III",
                "AM.mock.I", "AM.mock.III",
                "AM.WSN.I", "AM.WSN.II",
                "AM.WSN.PP1.I", "AM.WSN.PP1.II", "AM.WSN.PP1.III")
)

samples$CellType <- factor(samples$CellType, levels = c("AEC", "AM"))
samples$Treatment <- factor(samples$Treatment, levels = c("mock", "WSN", "WSN_PP1"))
samples$Replicate <- factor(samples$Replicate, levels = c("I", "II", "III"))

# Combined group label for plotting
samples$Group <- factor(paste(samples$CellType, samples$Treatment, sep = "."))

# Check sample names are exact match (should be TRUE)
all(rownames(samples) == colnames(rnaseqMatrix))
```

## Create DESeq2 object and filter low-count genes

```{r create-deseq-object}
# Create DESeq2 object
deseq2Data <- DESeqDataSetFromMatrix(countData = rnaseqMatrix, colData = samples,
                                     design = ~ Group)

# Determine how many genes pass the count filter
dim(deseq2Data)
dim(deseq2Data[rowSums(counts(deseq2Data)) > 10, ])

# Filter lowly expressed genes
deseq2Data <- deseq2Data[rowSums(counts(deseq2Data)) > 10, ]
```

# Diagnostic analyses

## Pairwise scatterplots

```{r scatterplots, eval=FALSE}
#*******************************************
# Diagnostic Analyses
#*******************************************

# Pairwise scatterplots of log2-transformed counts
# This can take a long time to run, so it is set to eval = FALSE.
pairs(log2(counts(deseq2Data) + 1), pch = 16)
```

## Run DESeq2 and rlog transformation

```{r run-deseq-rlog}
# Run DESeq2 pipeline
deseq2Data <- DESeq(deseq2Data)

# rlog transform counts
rld <- rlog(deseq2Data, blind = FALSE)
rlogcounts <- data.frame(assay(rld))
rownames(rlogcounts) <- rownames(deseq2Data)
```

## PCA: cleaned dataset

```{r pca-all-samples}
# PCA plot data
pcaData <- plotPCA(rld, intgroup = "Group", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
pcaData$Sample <- rownames(pcaData)

# PCA plot
# pdf("PCA_cleaned_samples.pdf")

ggplot(pcaData, aes(PC1, PC2, color = Group)) + geom_point(size = 2) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA: Cleaned Samples") +
  theme_bw(base_size = 14) + coord_fixed()

# dev.off()
```

## Correlation heatmap

```{r correlation-heatmap}
# Inspect rlog data using heatmap of pairwise correlation
cor_matrix <- cor(assay(rld))

# pdf("heatmap_rlog_correlation_cleaned_samples.pdf")

Heatmap(cor_matrix, name = "Correlation", column_title = "Sample Correlation: Cleaned Samples",
        row_title = "Samples", cluster_rows = TRUE, cluster_columns = TRUE)

# dev.off()
```

```{r qc-conclusion, eval=FALSE}
# QC conclusion:
# PCA and sample correlation analysis identified several samples that did not cluster
# with their expected treatment groups, suggesting potential technical variation.
# These outliers were removed prior to downstream analysis.
# The filtered dataset shows improved within-group clustering and overall correlation structure,
# supporting the use of the cleaned dataset for differential expression analysis.
```

# Split analysis by cell type

The dataset was analyzed separately by cell type so treatment effects could be evaluated within AEC and AM samples.

## AEC samples

```{r aec-analysis}
#*******************************************
# AEC Samples
#*******************************************

# Slice DESeq2 object for CellType = AEC
dds_AEC <- deseq2Data[, deseq2Data$CellType == "AEC"]
design(dds_AEC) <- ~ Treatment
dds_AEC <- DESeq(dds_AEC)

# rlog transform
rld_AEC <- rlog(dds_AEC, blind = FALSE)

# PCA plot data
pcaData_AEC <- plotPCA(rld_AEC, intgroup = "Treatment", returnData = TRUE)
percentVar_AEC <- round(100 * attr(pcaData_AEC, "percentVar"))
pcaData_AEC$Sample <- rownames(pcaData_AEC)

# PCA plot
#pdf("AEC_PCA_cleaned_samples.pdf")

ggplot(pcaData_AEC, aes(PC1, PC2, color = Treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_AEC[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_AEC[2], "% variance")) +
  ggtitle("AEC PCA: Cleaned Samples") + coord_fixed()

#dev.off()
```

## AM samples

```{r am-analysis}
#*******************************************
# AM Samples
#*******************************************

# Slice DESeq2 object for CellType = AM
dds_AM <- deseq2Data[, deseq2Data$CellType == "AM"]
design(dds_AM) <- ~ Treatment
dds_AM <- DESeq(dds_AM)

# rlog transform
rld_AM <- rlog(dds_AM, blind = FALSE)

# PCA plot data
pcaData_AM <- plotPCA(rld_AM, intgroup = "Treatment", returnData = TRUE)
percentVar_AM <- round(100 * attr(pcaData_AM, "percentVar"))
pcaData_AM$Sample <- rownames(pcaData_AM)

# PCA plot
# pdf("AM_PCA_cleaned_samples.pdf")

ggplot(pcaData_AM, aes(PC1, PC2, color = Treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_AM[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_AM[2], "% variance")) +
  ggtitle("AM PCA: Cleaned Samples") + coord_fixed()

# dev.off()
```

# Pairwise contrasts

```{r pairwise-contrasts}
#*******************************************
# Pairwise Contrasts
#*******************************************

# AEC: WSN vs mock
res_AEC_WSN_mock <- results(dds_AEC, contrast = c("Treatment", "WSN", "mock"))
resOrdered_AEC_WSN_mock <- res_AEC_WSN_mock[order(res_AEC_WSN_mock$padj, na.last = NA), ]

# AEC: WSN_PP1 vs WSN
res_AEC_WSNPP1_WSN <- results(dds_AEC, contrast = c("Treatment", "WSN_PP1", "WSN"))
resOrdered_AEC_WSNPP1_WSN <- res_AEC_WSNPP1_WSN[order(res_AEC_WSNPP1_WSN$padj, na.last = NA), ]

# AM: WSN vs mock
res_AM_WSN_mock <- results(dds_AM, contrast = c("Treatment", "WSN", "mock"))
resOrdered_AM_WSN_mock <- res_AM_WSN_mock[order(res_AM_WSN_mock$padj, na.last = NA), ]

# AM: WSN_PP1 vs WSN
res_AM_WSNPP1_WSN <- results(dds_AM, contrast = c("Treatment", "WSN_PP1", "WSN"))
resOrdered_AM_WSNPP1_WSN <- res_AM_WSNPP1_WSN[order(res_AM_WSNPP1_WSN$padj, na.last = NA), ]
```

# Differential expression summary

```{r de-summary}
#*******************************************
# Differential Expression Summary
#*******************************************

# Summary of differential expression results for each contrast
summary(res_AEC_WSN_mock)
summary(res_AEC_WSNPP1_WSN)
summary(res_AM_WSN_mock)
summary(res_AM_WSNPP1_WSN)

# Count DE genes with adjusted p-value < 0.005
DE_summary <- data.frame(Contrast = c("AEC: WSN vs mock",
                                      "AEC: WSN_PP1 vs WSN",
                                      "AM: WSN vs mock",
                                      "AM: WSN_PP1 vs WSN"),
                         
                         padj_0.005 = c(sum(res_AEC_WSN_mock$padj < 0.005, na.rm = TRUE),
                                        sum(res_AEC_WSNPP1_WSN$padj < 0.005, na.rm = TRUE),
                                        sum(res_AM_WSN_mock$padj < 0.005, na.rm = TRUE),
                                        sum(res_AM_WSNPP1_WSN$padj < 0.005, na.rm = TRUE)))
DE_summary
write.table(DE_summary, "DE_summary_counts_padj_0.005.txt", sep = "\t",
            row.names = FALSE, quote = FALSE)
```

# MA plots

```{r ma-plots}
#*******************************************
# MA Plots
#*******************************************

# AEC: WSN vs mock
plotMA(res_AEC_WSN_mock, main = "AEC: WSN vs mock")

# AEC: WSN_PP1 vs WSN
plotMA(res_AEC_WSNPP1_WSN, main = "AEC: WSN_PP1 vs WSN")

# AM: WSN vs mock
plotMA(res_AM_WSN_mock, main = "AM: WSN vs mock")

# AM: WSN_PP1 vs WSN
plotMA(res_AM_WSNPP1_WSN, main = "AM: WSN_PP1 vs WSN")
```

# Annotated result tables

```{r annotated-result-tables}
#*******************************************
# Annotated Result Tables
#*******************************************

# Helper function "annotate_results" to annotate and export DESeq2 results
annotate_results <- function(res, annotation_gene, filename)
  {results_df <- data.frame(GeneID = rownames(res), res)
  
  # Merge results with annotation file
  results_annot <- merge(results_df, annotation_gene,
                         by.x = "GeneID",
                         by.y = "Gene.stable.ID",
                         all.x = TRUE)
  
  # Sort annotated results by adjusted p-value, with NA values at the bottom
  results_annot <- results_annot[order(is.na(results_annot$padj), results_annot$padj),]
  
  write.table(results_annot, filename, sep = "\t", row.names = FALSE)
  return(results_annot)}

# AEC: WSN vs mock
results_AEC_WSN_mock_annot <- annotate_results(res_AEC_WSN_mock, annotation_gene,
                                               "AEC_WSN_vs_mock_DESeq2_annot.txt")

# AEC: WSN_PP1 vs WSN
results_AEC_WSNPP1_WSN_annot <- annotate_results(res_AEC_WSNPP1_WSN, annotation_gene,
                                                 "AEC_WSNPP1_vs_WSN_DESeq2_annot.txt")

# AM: WSN vs mock
results_AM_WSN_mock_annot <- annotate_results(res_AM_WSN_mock, annotation_gene,
                                              "AM_WSN_vs_mock_DESeq2_annot.txt")

# AM: WSN_PP1 vs WSN
results_AM_WSNPP1_WSN_annot <- annotate_results(res_AM_WSNPP1_WSN, annotation_gene,
                                                "AM_WSNPP1_vs_WSN_DESeq2_annot.txt")
```

# Volcano plots

```{r volcano-plots}
#*******************************************
# Volcano Plots
#*******************************************

# Gene symbols for AEC and AM
symbols_AEC <- mapIds(org.Mm.eg.db, keys = rownames(dds_AEC),
                      column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")

symbols_AM <- mapIds(org.Mm.eg.db, keys = rownames(dds_AM),
                     column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")


#*******************************************
# Volcano Plot: AEC WSN vs mock
#*******************************************

# Shrink log2 fold changes for volcano plot
res_AEC_WSN_mock_shrunk <- lfcShrink(dds_AEC, contrast = c("Treatment", "WSN", "mock"),
                                     res = res_AEC_WSN_mock, type = "normal")

# Use gene symbols as plot labels when available; otherwise keep Ensembl IDs
volcano_labels_AEC <- symbols_AEC[match(rownames(res_AEC_WSN_mock_shrunk), names(symbols_AEC))]
volcano_labels_AEC[is.na(volcano_labels_AEC)] <-
  rownames(res_AEC_WSN_mock_shrunk)[is.na(volcano_labels_AEC)]

# Select the top 25 genes by adjustedd p-value for labeling
top_genes_AEC <- head(rownames(res_AEC_WSN_mock_shrunk[order(res_AEC_WSN_mock_shrunk$padj,
                                                             na.last = NA), ]),25)

EnhancedVolcano(res_AEC_WSN_mock_shrunk, lab = volcano_labels_AEC,
                selectLab = volcano_labels_AEC[match(top_genes_AEC, rownames(res_AEC_WSN_mock_shrunk))],
                x = "log2FoldChange", y = "padj", title = "AEC: WSN vs mock",
                subtitle = "Infection response", pCutoff = 0.005, FCcutoff = 1.0,
                drawConnectors = TRUE, widthConnectors = 0.5,
                pointSize = 2.0, labSize = 3.0)

#*******************************************
# Volcano Plot: AEC WSN_PP1 vs WSN
#*******************************************

res_AEC_WSNPP1_WSN_shrunk <- lfcShrink(dds_AEC, contrast = c("Treatment", "WSN_PP1", "WSN"),
                                       res = res_AEC_WSNPP1_WSN, type = "normal")

volcano_labels_AEC <- symbols_AEC[match(rownames(res_AEC_WSNPP1_WSN_shrunk), names(symbols_AEC))]
volcano_labels_AEC[is.na(volcano_labels_AEC)] <-
  rownames(res_AEC_WSNPP1_WSN_shrunk)[is.na(volcano_labels_AEC)]

top_genes_AEC <- head(rownames(res_AEC_WSNPP1_WSN_shrunk[order(res_AEC_WSNPP1_WSN_shrunk$padj,
                                                               na.last = NA), ]), 25)

EnhancedVolcano(res_AEC_WSNPP1_WSN_shrunk, lab = volcano_labels_AEC,
                selectLab = volcano_labels_AEC[match(top_genes_AEC, rownames(res_AEC_WSNPP1_WSN_shrunk))],
                x = "log2FoldChange", y = "padj", title = "AEC: WSN_PP1 vs WSN",
                subtitle = "Inhibitor effect", pCutoff = 0.005, FCcutoff = 1.0,
                drawConnectors = TRUE, widthConnectors = 0.5,
                pointSize = 2.0, labSize = 3.0)

#*******************************************
# Volcano Plot: AM WSN vs mock
#*******************************************

res_AM_WSN_mock_shrunk <- lfcShrink(dds_AM, contrast = c("Treatment", "WSN", "mock"),
                                    res = res_AM_WSN_mock, type = "normal")

volcano_labels_AM <- symbols_AM[match(rownames(res_AM_WSN_mock_shrunk), names(symbols_AM))]
volcano_labels_AM[is.na(volcano_labels_AM)] <-
  rownames(res_AM_WSN_mock_shrunk)[is.na(volcano_labels_AM)]

top_genes_AM <- head(rownames(res_AM_WSN_mock_shrunk[order(res_AM_WSN_mock_shrunk$padj,
                                                           na.last = NA), ]), 25)

EnhancedVolcano(res_AM_WSN_mock_shrunk, lab = volcano_labels_AM,
                selectLab = volcano_labels_AM[match(top_genes_AM, rownames(res_AM_WSN_mock_shrunk))],
                x = "log2FoldChange", y = "padj", title = "AM: WSN vs mock",
                subtitle = "Infection response", pCutoff = 0.005, FCcutoff = 1.0,
                drawConnectors = TRUE, widthConnectors = 0.5,
                pointSize = 2.0, labSize = 3.0)

#*******************************************
# Volcano Plot: AM WSN_PP1 vs WSN
#*******************************************

res_AM_WSNPP1_WSN_shrunk <- lfcShrink(dds_AM, contrast = c("Treatment", "WSN_PP1", "WSN"),
                                      res = res_AM_WSNPP1_WSN, type = "normal")

volcano_labels_AM <- symbols_AM[match(rownames(res_AM_WSNPP1_WSN_shrunk), names(symbols_AM))]
volcano_labels_AM[is.na(volcano_labels_AM)] <-
  rownames(res_AM_WSNPP1_WSN_shrunk)[is.na(volcano_labels_AM)]

top_genes_AM <- head(rownames(res_AM_WSNPP1_WSN_shrunk[order(res_AM_WSNPP1_WSN_shrunk$padj,
                                                             na.last = NA), ]), 25)

EnhancedVolcano(res_AM_WSNPP1_WSN_shrunk, lab = volcano_labels_AM,
                selectLab = volcano_labels_AM[match(top_genes_AM, rownames(res_AM_WSNPP1_WSN_shrunk))],
                x = "log2FoldChange", y = "padj", title = "AM: WSN_PP1 vs WSN",
                subtitle = "Inhibitor effect", pCutoff = 0.005, FCcutoff = 1.0,
                drawConnectors = TRUE, widthConnectors = 0.5,
                pointSize = 2.0, labSize = 3.0)
```

# Heatmaps
```{r heatmaps}
#*******************************************
# Heatmaps
#*******************************************

# Helper functions

# Row z-score
cal_z_score <- function(x) {
  (x - mean(x)) / sd(x)}

# rlog-transformed counts for each cell type
rlogcounts_AEC <- data.frame(assay(rld_AEC))
rownames(rlogcounts_AEC) <- rownames(dds_AEC)

rlogcounts_AM <- data.frame(assay(rld_AM))
rownames(rlogcounts_AM) <- rownames(dds_AM)


#*******************************************
# Heatmap: AEC WSN vs mock
#*******************************************

# Candidate genes (top 25 by p-value)
top25_AEC_WSN_mock <- rownames(resOrdered_AEC_WSN_mock)[1:25]

candidates_AEC_WSN_mock <- data.frame(ID = top25_AEC_WSN_mock,
  Symbol = annotation_gene$Gene.name[match(top25_AEC_WSN_mock, annotation_gene$Gene.stable.ID)])

# Replace missing symbols with Ensembl IDs
candidates_AEC_WSN_mock$Symbol[is.na(candidates_AEC_WSN_mock$Symbol)] <-
  candidates_AEC_WSN_mock$ID[is.na(candidates_AEC_WSN_mock$Symbol)]

# Subset rlog data
table_hits_AEC_WSN_mock <- rlogcounts_AEC[candidates_AEC_WSN_mock$ID, ]

# Replace any NA values with 0
table_hits_AEC_WSN_mock[is.na(table_hits_AEC_WSN_mock)] <- 0

# Replace Ensembl IDs with gene symbols
rownames(table_hits_AEC_WSN_mock) <- candidates_AEC_WSN_mock$Symbol

# Calculate row z-scores
table_hits_AEC_WSN_mock_norm <- t(apply(table_hits_AEC_WSN_mock, 1, cal_z_score))

# Plot heatmap to pdf
#pdf("AEC_WSN_vs_mock_heatmap_top25.pdf")
Heatmap(table_hits_AEC_WSN_mock_norm, name = "Row Z-Score", column_title = "AEC: WSN vs mock",
        row_title = "Genes", cluster_rows = TRUE, cluster_columns = TRUE)
#dev.off()

#*******************************************
# Heatmap: AEC WSN_PP1 vs WSN
#*******************************************

# Candidate genes (top 25 by p-value)
top25_AEC_WSNPP1_WSN <- rownames(resOrdered_AEC_WSNPP1_WSN)[1:25]

candidates_AEC_WSNPP1_WSN <- data.frame(ID = top25_AEC_WSNPP1_WSN,
  Symbol = annotation_gene$Gene.name[match(top25_AEC_WSNPP1_WSN, annotation_gene$Gene.stable.ID)])

# Replace missing symbols with Ensembl IDs
candidates_AEC_WSNPP1_WSN$Symbol[is.na(candidates_AEC_WSNPP1_WSN$Symbol)] <-
  candidates_AEC_WSNPP1_WSN$ID[is.na(candidates_AEC_WSNPP1_WSN$Symbol)]

# Subset rlog data
table_hits_AEC_WSNPP1_WSN <- rlogcounts_AEC[candidates_AEC_WSNPP1_WSN$ID, ]

# Replace any NA values with 0
table_hits_AEC_WSNPP1_WSN[is.na(table_hits_AEC_WSNPP1_WSN)] <- 0

# Replace Ensembl IDs with gene symbols
rownames(table_hits_AEC_WSNPP1_WSN) <- candidates_AEC_WSNPP1_WSN$Symbol

# Calculate row z-scores
table_hits_AEC_WSNPP1_WSN_norm <- t(apply(table_hits_AEC_WSNPP1_WSN, 1, cal_z_score))

# Plot heatmap
# pdf("AEC_WSNPP1_vs_WSN_heatmap_top25.pdf")
Heatmap(table_hits_AEC_WSNPP1_WSN_norm, name = "Row Z-Score", column_title = "AEC: WSN_PP1 vs WSN",
        row_title = "Genes", cluster_rows = TRUE, cluster_columns = TRUE)
# dev.off()

#*******************************************
# Heatmap: AM WSN vs mock
#*******************************************

# Candidate genes (top 25 by p-value)
top25_AM_WSN_mock <- rownames(resOrdered_AM_WSN_mock)[1:25]

candidates_AM_WSN_mock <- data.frame(ID = top25_AM_WSN_mock,
  Symbol = annotation_gene$Gene.name[match(top25_AM_WSN_mock, annotation_gene$Gene.stable.ID)])

# Replace missing symbols with Ensembl IDs
candidates_AM_WSN_mock$Symbol[is.na(candidates_AM_WSN_mock$Symbol)] <-
  candidates_AM_WSN_mock$ID[is.na(candidates_AM_WSN_mock$Symbol)]

# Subset rlog data
table_hits_AM_WSN_mock <- rlogcounts_AM[candidates_AM_WSN_mock$ID, ]

# Replace any NA values with 0
table_hits_AM_WSN_mock[is.na(table_hits_AM_WSN_mock)] <- 0

# Replace Ensembl IDs with gene symbols
rownames(table_hits_AM_WSN_mock) <- candidates_AM_WSN_mock$Symbol

# Calculate row z-scores
table_hits_AM_WSN_mock_norm <- t(apply(table_hits_AM_WSN_mock, 1, cal_z_score))

# Plot heatmap
#pdf("AM_WSN_vs_mock_heatmap_top25.pdf")
Heatmap(table_hits_AM_WSN_mock_norm, name = "Row Z-Score", column_title = "AM: WSN vs mock",
        row_title = "Genes", cluster_rows = TRUE, cluster_columns = TRUE)
#dev.off()

#*******************************************
# Heatmap: AM WSN_PP1 vs WSN
#*******************************************

# Candidate genes (top 25 by p-value)
top25_AM_WSNPP1_WSN <- rownames(resOrdered_AM_WSNPP1_WSN)[1:25]

candidates_AM_WSNPP1_WSN <- data.frame(ID = top25_AM_WSNPP1_WSN,
  Symbol = annotation_gene$Gene.name[match(top25_AM_WSNPP1_WSN, annotation_gene$Gene.stable.ID)])

# Replace missing symbols with Ensembl IDs
candidates_AM_WSNPP1_WSN$Symbol[is.na(candidates_AM_WSNPP1_WSN$Symbol)] <-
  candidates_AM_WSNPP1_WSN$ID[is.na(candidates_AM_WSNPP1_WSN$Symbol)]

# Subset rlog data
table_hits_AM_WSNPP1_WSN <- rlogcounts_AM[candidates_AM_WSNPP1_WSN$ID, ]

# Replace any NA values with 0
table_hits_AM_WSNPP1_WSN[is.na(table_hits_AM_WSNPP1_WSN)] <- 0

# Replace Ensembl IDs with gene symbols
rownames(table_hits_AM_WSNPP1_WSN) <- candidates_AM_WSNPP1_WSN$Symbol

# Calculate row z-scores
table_hits_AM_WSNPP1_WSN_norm <- t(apply(table_hits_AM_WSNPP1_WSN, 1, cal_z_score))

# Plot heatmap
# pdf("AM_WSNPP1_vs_WSN_heatmap_top25.pdf")
Heatmap(table_hits_AM_WSNPP1_WSN_norm, name = "Row Z-Score", column_title = "AM: WSN_PP1 vs WSN",
        row_title = "Genes", cluster_rows = TRUE, cluster_columns = TRUE)
#dev.off()

# Show heatmap in Rmd
Heatmap(table_hits_AM_WSNPP1_WSN_norm, name = "Row Z-Score", column_title = "AM: WSN_PP1 vs WSN",
        row_title = "Genes", cluster_rows = TRUE, cluster_columns = TRUE)

```

```{r session-info}
#*******************************************
# End of Script
#*******************************************
sessionInfo()
```
