---
title: "RNA-seq DESeq2 Differential Expression and QC Workflow"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    latex_engine: xelatex
fontsize: 10pt
geometry: margin=0.8in
mainfont: Arial
header-includes:
  - \usepackage{titling}
  - \usepackage{graphicx}
  - \usepackage{xcolor}
  - \usepackage{sectsty}
  - \definecolor{brandnavy}{HTML}{1F3347}
  - \definecolor{brandteal}{HTML}{2C8C99}
  - \sectionfont{\color{brandnavy}}
  - \subsectionfont{\color{brandnavy}}
  - \pretitle{\begin{center}\includegraphics[width=0.30\textwidth]{logo.png}\\[0.8em]\Large\bfseries\color{brandnavy}}
  - \posttitle{\end{center}}
  - \predate{\begin{center}\large\color{brandnavy}}
  - \postdate{\end{center}}
  - \newpage
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                        fig.align = "center", out.width = "85%")


set.seed(42)
```

# Overview

This workflow demonstrates a reproducible RNA-seq differential expression and 
reporting pipeline using the public GEO dataset GSE203539.

Raw sequencing data were originally processed through a Galaxy-based RNA-seq 
pipeline including quality assessment, read trimming, alignment to the GRCm39 
reference genome, and feature-level quantification prior to downstream analysis.

This repository focuses on downstream quality control, outlier sensitivity 
assessment, DESeq2 differential expression analysis, and structured result export
workflows in R.

# Analysis Plan

1. Load RNA-seq count data from `data/all_counts.txt`
2. Parse sample metadata
3. Perform QC and exploratory diagnostics on all samples
4. Evaluate pre-identified outlier samples through sensitivity analysis
5. Run filtered DESeq2 differential expression analysis
6. Export QC figures, DEG tables, and summary outputs to `results/`
7. Define high-confidence DEGs using adjusted p-value ≤ 0.005 and absolute log2 fold change > 2.

# Load packages {.unnumbered .unlisted}

```{r load-packages}
required_packages <- c(
  "DESeq2",
  "EnhancedVolcano",
  "org.Mm.eg.db",
  "AnnotationDbi",
  "SummarizedExperiment",
  "ComplexHeatmap",
  "circlize",
  "grid",
  "ggplot2",
  "ggrepel",
  "dplyr",
  "tidyr",
  "readr",
  "stringr",
  "tibble",
  "purrr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace,
                                              logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required package(s): ",
    paste(missing_packages, collapse = ", "),
    "\nInstall missing packages before knitting this workflow."
  )
}

library(DESeq2)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(SummarizedExperiment)
library(org.Mm.eg.db)
library(AnnotationDbi)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(tibble)
library(purrr)
```

# Project paths {.unnumbered .unlisted}

```{r project-paths}
data_dir <- "data"
results_dir <- "results"
figures_dir <- file.path(results_dir, "figures")
tables_dir <- file.path(results_dir, "tables")
deg_dir <- file.path(results_dir, "deg_outputs")
qc_dir <- file.path(results_dir, "qc")

walk(
  c(results_dir, figures_dir, tables_dir, deg_dir, qc_dir),
  ~ dir.create(.x, recursive = TRUE, showWarnings = FALSE)
)

counts_file <- file.path(data_dir, "all_counts.txt")
annotation_file <- file.path(data_dir, "GRCm39_annotation.txt")

if (!file.exists(counts_file)) {
  stop("Missing required count matrix: ", counts_file)
}

# Portfolio-aligned color palette
brand_navy <- "#1F3347"
brand_teal <- "#2C8C99"
brand_red <- "#A94E4E"
brand_green <- "#3F7F5F"
brand_blue <- "#3F6FA8"
brand_purple <- "#6F5A9A"
brand_gold <- "#9A7A2F"
brand_gray <- "#4B5563"
brand_light_gray <- "#EEF1F4"

group_colors <- c(
  AEC.mock = brand_red,
  AEC.WSN = brand_green,
  AEC.WSN_PP1 = brand_blue,
  AM.mock = brand_purple,
  AM.WSN = brand_gold,
  AM.WSN_PP1 = brand_teal
)

condition_colors <- c(
  mock = brand_red,
  WSN = brand_green,
  WSN_PP1 = brand_blue
)

cell_type_colors <- c(
  AEC = brand_teal,
  AM = brand_navy
)

outlier_colors <- c(
  `FALSE` = brand_light_gray,
  `TRUE` = brand_teal
)

direction_colors <- c(
  strict_down = brand_blue,
  strict_up = brand_red
)

sensitivity_colors <- c(
  not_strict_in_either = brand_gray,
  strict_all_only = brand_gold,
  strict_filtered_only = brand_teal,
  strict_in_both = brand_purple
)

distance_colors <- circlize::colorRamp2(
  c(0, 200, 400),
  c(brand_navy, brand_light_gray, brand_red)
)

zscore_colors <- circlize::colorRamp2(
  c(-2, 0, 2),
  c(brand_blue, "white", brand_red)
)
```

# Import count matrix {.unnumbered .unlisted}

```{r import-counts, results = 'hide'}
raw_counts <- read.delim(
  counts_file,
  header = TRUE,
  sep = "\t",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

if (ncol(raw_counts) < 3) {
  stop("Count matrix appears too small. Expected one gene ID column plus sample count columns.")
}

# Support both simple count matrices and featureCounts-style tables.
featurecounts_annotation_cols <- c("Chr", "Start", "End", "Strand", "Length")
possible_gene_cols <- c("gene_id", "Geneid", "GeneID", "Gene.stable.ID", "gene", "id")

gene_col <- intersect(possible_gene_cols, names(raw_counts))[1]

if (is.na(gene_col)) {
  gene_col <- names(raw_counts)[1]
  message("No standard gene ID column found. Using first column as gene_id: ", gene_col)
}

counts_tbl <- raw_counts %>%
  rename(gene_id = all_of(gene_col)) %>%
  select(-any_of(featurecounts_annotation_cols))

# If featureCounts used full BAM paths as sample names, simplify them where possible.
clean_sample_name <- function(x) {
  x %>%
    basename() %>%
    str_remove("\\.bam$") %>%
    str_remove("\\.sam$") %>%
    str_remove("\\.sorted$")
}

names(counts_tbl) <- c("gene_id", clean_sample_name(names(counts_tbl)[-1]))

sample_cols <- setdiff(names(counts_tbl), "gene_id")

counts_tbl <- counts_tbl %>%
  mutate(across(all_of(sample_cols), as.numeric))

if (any(is.na(counts_tbl[sample_cols]))) {
  stop("Count matrix contains NA values after numeric conversion. Check input formatting.")
}

count_matrix <- counts_tbl %>%
  group_by(gene_id) %>%
  summarise(across(all_of(sample_cols), ~ sum(.x, na.rm = TRUE)), .groups = "drop") %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

storage.mode(count_matrix) <- "integer"

if (any(count_matrix < 0)) {
  stop("Count matrix contains negative values, which are invalid for DESeq2.")
}

cat("Genes:", nrow(count_matrix), "\n")
cat("Samples:", ncol(count_matrix), "\n")
```

# Build sample metadata {.unnumbered .unlisted}

```{r build-sample-metadata, results='hide'}
outlier_samples <- c("AEC.mock.III", "AEC.WSN.PP1.I", "AM.mock.II", "AM.WSN.III")

metadata <- tibble(sample_id = colnames(count_matrix)) %>%
  mutate(
    cell_type = case_when(
      str_starts(sample_id, "AEC\\.") ~ "AEC",
      str_starts(sample_id, "AM\\.") ~ "AM",
      TRUE ~ NA_character_
    ),
    condition = case_when(
      str_detect(sample_id, "\\.mock\\.") ~ "mock",
      str_detect(sample_id, "\\.WSN\\.PP1\\.") ~ "WSN_PP1",
      str_detect(sample_id, "\\.WSN\\.PP\\.") ~ "WSN_PP1",
      str_detect(sample_id, "\\.WSN\\.") ~ "WSN",
      TRUE ~ NA_character_
    ),
    replicate = str_extract(sample_id, "[^.]+$"),
    outlier_flag = sample_id %in% outlier_samples
  ) %>%
  mutate(
    cell_type = factor(cell_type, levels = c("AEC", "AM")),
    condition = factor(condition, levels = c("mock", "WSN", "WSN_PP1")),
    replicate = factor(replicate),
    group = factor(paste(cell_type, condition, sep = "."))
  )

if (any(is.na(metadata$cell_type)) || any(is.na(metadata$condition))) {
  print(metadata)
  stop("Unable to infer cell_type or condition for one or more samples.
       Check sample naming conventions.")
}

missing_outliers <- setdiff(outlier_samples, metadata$sample_id)

if (length(missing_outliers) > 0) {
  warning("Pre-specified outlier sample(s) not found in count matrix: ",
          paste(missing_outliers, collapse = ", "))
}

if (!all(metadata$sample_id == colnames(count_matrix))) {
  stop("Metadata sample order does not match count matrix column order.")
}

write_csv(metadata, file.path(qc_dir, "sample_metadata_with_outlier_flags.csv"))
```

# Optional gene annotation {.unnumbered .unlisted}

```{r read-annotation}
annotation_file <- file.path("data", "GRCm39_annotation.txt")

if (file.exists(annotation_file)) {
  annotation <- read.delim(
    annotation_file,
    header = TRUE,
    sep = "\t",
    quote = "\"",
    check.names = TRUE,
    stringsAsFactors = FALSE
  )

  if ("Gene.stable.ID" %in% names(annotation)) {
    annotation_gene <- annotation[!duplicated(annotation$Gene.stable.ID), ]
  } else {
    warning("Annotation file found, but Gene.stable.ID column was not detected.
            Results will be exported without annotation merge.")
    annotation_gene <- NULL
  }
} else {
  warning("Optional annotation file not found: ", annotation_file, ". 
          Results will be exported without annotation merge.")
  annotation_gene <- NULL
}
```

# QC on all samples

Quality control is performed before excluding any samples. The outlier samples 
are flagged visually and retained in this section.

## Library size

```{r qc-library-size-all}
library_sizes <- metadata %>%
  mutate(total_counts = colSums(count_matrix)[sample_id])

p_library <- ggplot(
  library_sizes,
  aes(
    x = reorder(sample_id, total_counts),
    y = total_counts,
    fill = group
  )
) +
  geom_col(color = "grey35", linewidth = 0.2) +
  geom_col(
    data = library_sizes %>% filter(outlier_flag),
    aes(x = reorder(sample_id, total_counts), y = total_counts, fill = group),
    color = "black",
    linewidth = 0.7
  ) +
  scale_fill_manual(values = group_colors) +
  coord_flip() +
  facet_wrap(~ cell_type, scales = "free_y") +
  labs(
    title = "Library size by sample: all samples",
    x = "Sample",
    y = "Total counts",
    fill = "Group"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = brand_navy),
    panel.grid.minor = element_blank()
  )

p_library

ggsave(
  file.path(figures_dir, "qc_library_size_all_samples.png"),
  p_library,
  width = 9,
  height = 6,
  dpi = 300
)

write_csv(library_sizes, file.path(qc_dir, "library_sizes_all_samples.csv"))
```

<!-- Variance-stabilized count transformation is performed prior to PCA and sample-distance analysis. -->

```{r qc-vst-all}
keep_genes_qc <- rowSums(count_matrix >= 10) >= 2

if (sum(keep_genes_qc) < 100) {
  stop("Too few genes pass the QC expression filter. Check count matrix input.")
}

dds_qc_all <- DESeqDataSetFromMatrix(
  countData = count_matrix[keep_genes_qc, ],
  colData = metadata %>% as.data.frame() %>% column_to_rownames("sample_id"),
  design = ~ cell_type + condition
)

vsd_qc_all <- vst(dds_qc_all, blind = TRUE)
```

## PCA: all samples 

```{r pca-all-samples, fig.width=7, fig.height=3.8, out.width="85%"}
pca_all <- plotPCA(
  vsd_qc_all,
  intgroup = "group",
  returnData = TRUE
) %>%
  rownames_to_column("sample_id") %>%
  dplyr::select(sample_id, PC1, PC2, group) %>%
  dplyr::left_join(
    metadata %>%
      dplyr::select(sample_id, outlier_flag),
    by = "sample_id"
  ) %>%
  dplyr::mutate(
    outlier_label = ifelse(outlier_flag, sample_id, NA_character_)
  )

percent_var_all <- round(100 * attr(pca_all, "percentVar"))

p_pca_all <- ggplot(
  pca_all,
  aes(
    x = PC1,
    y = PC2,
    color = group
  )
) +
  geom_point(size = 3, alpha = 0.95) +
  ggrepel::geom_text_repel(
    aes(label = outlier_label),
    size = 3,
    max.overlaps = Inf,
    box.padding = 0.35,
    point.padding = 0.25,
    min.segment.length = 0,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "PCA of variance-stabilized counts: all samples",
    x = paste0("PC1: ", percent_var_all[1], "% variance"),
    y = paste0("PC2: ", percent_var_all[2], "% variance"),
    color = "Group"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = brand_navy),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

p_pca_all

ggsave(
  file.path(figures_dir, "qc_pca_all_samples.png"),
  p_pca_all,
  width = 7,
  height = 3.8,
  dpi = 300
)

write_csv(pca_all, file.path(qc_dir, "pca_all_samples.csv"))
```

## Sample distance heatmap: all samples

```{r qc-distance-all}
sample_dist_all <- dist(t(assay(vsd_qc_all)))
sample_dist_mat_all <- as.matrix(sample_dist_all)

rownames(sample_dist_mat_all) <- colnames(vsd_qc_all)
colnames(sample_dist_mat_all) <- colnames(vsd_qc_all)

annotation_col_all <- metadata %>%
  dplyr::select(sample_id, cell_type, condition, outlier_flag) %>%
  dplyr::mutate(
    cell_type = factor(cell_type),
    condition = factor(condition),
    outlier_flag = factor(outlier_flag, levels = c(FALSE, TRUE))
  ) %>%
  as.data.frame()

rownames(annotation_col_all) <- annotation_col_all$sample_id
annotation_col_all$sample_id <- NULL

ha_all <- ComplexHeatmap::HeatmapAnnotation(
  df = annotation_col_all,
  which = "column",
  col = list(
    cell_type = cell_type_colors,
    condition = condition_colors,
    outlier_flag = outlier_colors
  )
)

ht_all <- ComplexHeatmap::Heatmap(
  sample_dist_mat_all,
  name = "Distance",
  col = distance_colors,
  top_annotation = ha_all,
  column_title = "All samples",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = grid::gpar(fontsize = 7),
  column_names_gp = grid::gpar(fontsize = 7),
  column_names_rot = 45
)

png(file.path(figures_dir, "qc_sample_distance_all_samples.png"), width = 2600, height = 2200, res = 300)
ComplexHeatmap::draw(ht_all, heatmap_legend_side = "right", annotation_legend_side = "right")
invisible(dev.off())

ComplexHeatmap::draw(ht_all, heatmap_legend_side = "right", annotation_legend_side = "right")
```

# Helper functions {.unnumbered .unlisted}

```{r helper-functions}
make_dds_for_cell <- function(counts, sample_metadata, cell, remove_outliers = FALSE) {
  meta <- sample_metadata %>%
    filter(cell_type == cell)

  if (remove_outliers) {
    meta <- meta %>% filter(!outlier_flag)
  }

  mat <- counts[, meta$sample_id, drop = FALSE]
  keep <- rowSums(mat >= 10) >= 2
  mat <- mat[keep, , drop = FALSE]

  meta <- meta %>%
    mutate(condition = droplevels(condition))

  if (nlevels(meta$condition) < 2) {
    stop("Fewer than two treatment conditions available for ", cell, ".")
  }

  dds <- DESeqDataSetFromMatrix(
    countData = mat,
    colData = meta %>% as.data.frame() %>% column_to_rownames("sample_id"),
    design = ~ condition
  )

  DESeq(dds)
}

extract_contrast <- function(dds, cell, contrast_label, numerator, denominator, alpha = 0.05) {
  available_conditions <- unique(as.character(colData(dds)$condition))

  if (!all(c(numerator, denominator) %in% available_conditions)) {
    warning("Skipping ", contrast_label, ": missing required condition.")
    return(NULL)
  }

  res <- results(
    dds,
    contrast = c("condition", numerator, denominator),
    alpha = alpha
  )

  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    mutate(
      cell_type = cell,
      contrast = contrast_label,
      numerator = numerator,
      denominator = denominator,
      strict_deg = !is.na(padj) & padj <= 0.005 & abs(log2FoldChange) > 2,
      regulation = case_when(
        strict_deg & log2FoldChange > 0 ~ "up",
        strict_deg & log2FoldChange < 0 ~ "down",
        TRUE ~ "not_strict"
      )
    ) %>%
    relocate(cell_type, contrast, gene_id)
}

run_cell_analysis <- function(counts, sample_metadata, cell, analysis_set, remove_outliers = FALSE) {
  dds <- make_dds_for_cell(counts, sample_metadata, cell, remove_outliers = remove_outliers)

  comparisons <- tibble(
    contrast = c(
      paste0(cell, "_WSN_vs_mock"),
      paste0(cell, "_WSN_PP1_vs_WSN")
    ),
    numerator = c("WSN", "WSN_PP1"),
    denominator = c("mock", "WSN")
  )

  results_tbl <- pmap_dfr(
    comparisons,
    ~ extract_contrast(dds, cell = cell, contrast_label = ..1, numerator = ..2, denominator = ..3)
  ) %>%
    mutate(analysis_set = analysis_set) %>%
    relocate(analysis_set, .after = contrast)

  list(dds = dds, results = results_tbl)
}

summarize_degs <- function(results_tbl) {
  results_tbl %>%
    group_by(analysis_set, cell_type, contrast) %>%
    summarise(
      genes_tested = n(),
      padj_0_05 = sum(!is.na(padj) & padj <= 0.05),
      padj_0_005 = sum(!is.na(padj) & padj <= 0.005),
      strict_degs = sum(strict_deg, na.rm = TRUE),
      strict_up = sum(strict_deg & log2FoldChange > 0, na.rm = TRUE),
      strict_down = sum(strict_deg & log2FoldChange < 0, na.rm = TRUE),
      .groups = "drop"
    )
}

annotate_results <- function(results_tbl, annotation_gene) {
  if (is.null(annotation_gene)) {
    return(results_tbl)
  }

  results_tbl %>%
    left_join(annotation_gene, by = c("gene_id" = "Gene.stable.ID"))
}
```

# Sensitivity analysis
## All-sample vs filtered DESeq2 results

This section evaluates whether the pre-specified outlier handling decision alters 
differential expression results. Both all-sample and filtered analyses are retained 
and exported for transparency.

High-confidence DEGs were defined using an adjusted p-value threshold ≤ 0.005
and an absolute log2 fold-change > 2. These thresholds were used for DEG summaries, 
sensitivity comparisons, and downstream reporting exports.

```{r sensitivity-analysis, results='hide'}
all_sample_analyses <- map(
  c("AEC", "AM"),
  ~ run_cell_analysis(count_matrix, metadata, cell = .x, analysis_set = "all_samples",
                      remove_outliers = FALSE)
)

filtered_analyses <- map(
  c("AEC", "AM"),
  ~ run_cell_analysis(count_matrix, metadata, cell = .x, analysis_set = "filtered",
                      remove_outliers = TRUE)
)

all_sample_results <- map_dfr(all_sample_analyses, "results")
filtered_results <- map_dfr(filtered_analyses, "results")

sensitivity_results <- bind_rows(all_sample_results, filtered_results)
sensitivity_summary <- summarize_degs(sensitivity_results)

write_csv(sensitivity_results, file.path(deg_dir, "deseq2_results_all_vs_filtered.csv"))
write_csv(sensitivity_summary, file.path(tables_dir, "deg_summary_all_vs_filtered.csv"))

sensitivity_summary
```

```{r sensitivity-gene-level-comparison, results = 'hide'}
sensitivity_gene_comparison <- sensitivity_results %>%
  select(cell_type, contrast, analysis_set, gene_id, log2FoldChange, padj, strict_deg) %>%
  pivot_wider(
    names_from = analysis_set,
    values_from = c(log2FoldChange, padj, strict_deg),
    names_sep = "__"
  ) %>%
  mutate(
    strict_deg__all_samples = replace_na(strict_deg__all_samples, FALSE),
    strict_deg__filtered = replace_na(strict_deg__filtered, FALSE),
    delta_log2fc = log2FoldChange__filtered - log2FoldChange__all_samples,
    strict_status = case_when(
      strict_deg__all_samples & strict_deg__filtered ~ "strict_in_both",
      strict_deg__all_samples & !strict_deg__filtered ~ "strict_all_only",
      !strict_deg__all_samples & strict_deg__filtered ~ "strict_filtered_only",
      TRUE ~ "not_strict_in_either"
    )
  )

sensitivity_overlap_summary <- sensitivity_gene_comparison %>%
  group_by(cell_type, contrast, strict_status) %>%
  summarise(n_genes = n(), .groups = "drop")

write_csv(sensitivity_gene_comparison, file.path(tables_dir, "sensitivity_gene_level_comparison.csv"))
write_csv(sensitivity_overlap_summary, file.path(tables_dir, "sensitivity_overlap_summary.csv"))

sensitivity_overlap_summary
```

```{r sensitivity-plot}
p_sensitivity <- sensitivity_gene_comparison %>%
  filter(!is.na(log2FoldChange__all_samples), !is.na(log2FoldChange__filtered)) %>%
  ggplot(aes(x = log2FoldChange__all_samples, y = log2FoldChange__filtered, color = strict_status)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ contrast, scales = "free") +
  labs(
    title = "Sensitivity analysis: all-sample vs filtered log2 fold change",
    x = "All-sample log2 fold change",
    y = "Filtered log2 fold change",
    color = "Strict DEG status"
  ) +
  theme_bw(base_size = 12)

p_sensitivity

ggsave(file.path(figures_dir, "sensitivity_log2fc_all_vs_filtered.png"), p_sensitivity,
       width = 10, height = 7, dpi = 300)
```
The sensitivity analysis shows that most high-confidence DEG calls are stable across
filtering decisions, while selected contrasts, especially AM inhibitor-response comparisons,
show larger shifts after outlier removal. This supports retaining both all-sample
and filtered outputs for transparency.

# Filtered DESeq2 workflow

The remaining workflow uses the filtered dataset, excluding the explicitly flagged outlier samples.

```{r filtered-metadata}
metadata_filtered <- metadata %>%
  filter(!outlier_flag)

count_matrix_filtered <- count_matrix[, metadata_filtered$sample_id, drop = FALSE]

write_csv(metadata_filtered, file.path(qc_dir, "sample_metadata_filtered.csv"))
```

## Filtered PCA

```{r filtered-vst-pca, fig.width=7, fig.height=2.8}
metadata_filtered <- metadata %>%
  dplyr::filter(!outlier_flag)

count_matrix_filtered <- count_matrix[, metadata_filtered$sample_id, drop = FALSE]

keep_genes_filtered <- rowSums(count_matrix_filtered >= 10) >= 2

dds_qc_filtered <- DESeq2::DESeqDataSetFromMatrix(
  countData = count_matrix_filtered[keep_genes_filtered, ],
  colData = metadata_filtered %>%
    as.data.frame() %>%
    tibble::column_to_rownames("sample_id"),
  design = ~ cell_type + condition
)

vsd_qc_filtered <- DESeq2::vst(dds_qc_filtered, blind = TRUE)

pca_filtered <- DESeq2::plotPCA(
  vsd_qc_filtered,
  intgroup = "group",
  returnData = TRUE
) %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::select(sample_id, PC1, PC2, group)

percent_var_filtered <- round(100 * attr(pca_filtered, "percentVar"))

p_pca_filtered <- ggplot2::ggplot(
  pca_filtered,
  ggplot2::aes(
    x = PC1,
    y = PC2,
    color = group
  )
) +
  ggplot2::geom_point(size = 2, alpha = 0.95) +
  ggplot2::scale_color_manual(values = group_colors) +
  ggplot2::labs(
    title = "PCA of variance-stabilized counts: filtered samples",
    x = paste0("PC1: ", percent_var_filtered[1], "% variance"),
    y = paste0("PC2: ", percent_var_filtered[2], "% variance"),
    color = "Group"
  ) +
  ggplot2::coord_fixed(ratio = 1) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 12, color = brand_navy),
    legend.position = "right",
    panel.grid.minor = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(4, 4, 4, 4)
  )

p_pca_filtered

ggplot2::ggsave(
  file.path(figures_dir, "qc_pca_filtered_samples.png"),
  p_pca_filtered,
  width = 7,
  height = 2.8,
  dpi = 300
)

readr::write_csv(pca_filtered, file.path(qc_dir, "pca_filtered_samples.csv"))
```
PCA of the filtered dataset showed strong separation by cell type along PC1 (93% of variance),
while condition-related differences were captured along PC2 (3% of variance). These patterns
support analyzing AEC and AM samples as separate pairwise condition comparisons.

# Export filtered DEG results {.unnumbered .unlisted}

```{r filtered-result-exports, results = 'hide'}
filtered_results_annotated <- annotate_results(filtered_results, annotation_gene)
filtered_summary <- summarize_degs(filtered_results)
strict_filtered_results <- filtered_results_annotated %>%
  filter(strict_deg) %>%
  arrange(cell_type, contrast, padj, desc(abs(log2FoldChange)))

write_csv(filtered_results_annotated, file.path(deg_dir, "deseq2_results_filtered_annotated.csv"))
write_csv(strict_filtered_results, file.path(deg_dir, "strict_deg_results_filtered_padj005_log2fc2.csv"))
write_csv(filtered_summary, file.path(tables_dir, "deg_summary_filtered.csv"))

# Export one annotated table per contrast for downstream Python reporting.
filtered_results_annotated %>%
  group_split(contrast) %>%
  walk(function(tbl) {
    contrast_name <- unique(tbl$contrast)
    output_file <- file.path(deg_dir, paste0(contrast_name, "_DESeq2_annot.csv"))
    write_csv(tbl, output_file)
  })

filtered_summary
```

# Filtered DEG visualizations

## DEG count summary

```{r deg-count-plot}
p_deg_counts <- filtered_summary %>%
  select(cell_type, contrast, strict_up, strict_down) %>%
  pivot_longer(
    cols = c(strict_up, strict_down),
    names_to = "direction",
    values_to = "n_genes"
  ) %>%
  ggplot(aes(x = contrast, y = n_genes, fill = direction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  labs(
    title = "Strict DEG counts in filtered analysis",
    x = "Contrast",
    y = "Number of genes",
    fill = "Direction"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = brand_navy),
    panel.grid.minor = element_blank()
  )

p_deg_counts

ggsave(
  file.path(figures_dir, "strict_deg_counts_filtered.png"),
  p_deg_counts,
  width = 8,
  height = 5,
  dpi = 300
)
```

## EnhancedVolcano plots with gene labels

```{r enhanced-volcano-plots, results='markup', fig.show='hold', fig.width=5.2, fig.height=4.2, out.width='49%'}
make_enhanced_volcano <- function(dds, results_tbl, contrast_id, numerator, denominator) {
  res <- DESeq2::results(
    dds,
    contrast = c("condition", numerator, denominator)
  )

  res_shrunk <- DESeq2::lfcShrink(
    dds,
    contrast = c("condition", numerator, denominator),
    res = res,
    type = "normal"
  )

  symbols <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys = rownames(res_shrunk),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  labels <- symbols[match(rownames(res_shrunk), names(symbols))]
  labels[is.na(labels)] <- rownames(res_shrunk)[is.na(labels)]

  top_genes <- results_tbl %>%
    dplyr::filter(contrast == contrast_id) %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = 15) %>%
    dplyr::pull(gene_id)

  EnhancedVolcano::EnhancedVolcano(
    res_shrunk,
    lab = labels,
    selectLab = labels[match(top_genes, rownames(res_shrunk))],
    x = "log2FoldChange",
    y = "padj",
    title = contrast_id,
    subtitle = paste(numerator, "vs", denominator),
    pCutoff = 0.005,
    FCcutoff = 2,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    pointSize = 2.3,
    labSize = 3.4,
    titleLabSize = 12,
    subtitleLabSize = 10,
    axisLabSize = 10,
    legendLabSize = 8,
    legendIconSize = 3.0
  )
}

filtered_dds_AEC <- filtered_analyses[[1]]$dds
filtered_dds_AM <- filtered_analyses[[2]]$dds

volcano_specs <- tibble::tibble(
  cell_type = c("AEC", "AEC", "AM", "AM"),
  contrast = c(
    "AEC_WSN_vs_mock",
    "AEC_WSN_PP1_vs_WSN",
    "AM_WSN_vs_mock",
    "AM_WSN_PP1_vs_WSN"
  ),
  numerator = c("WSN", "WSN_PP1", "WSN", "WSN_PP1"),
  denominator = c("mock", "WSN", "mock", "WSN")
)

for (i in seq_len(nrow(volcano_specs))) {
  spec <- volcano_specs[i, ]
  dds <- if (spec$cell_type == "AEC") filtered_dds_AEC else filtered_dds_AM

  volcano_plot <- make_enhanced_volcano(
    dds = dds,
    results_tbl = filtered_results,
    contrast_id = spec$contrast,
    numerator = spec$numerator,
    denominator = spec$denominator
  )

  png(
    file.path(figures_dir, paste0("enhanced_volcano_", spec$contrast, ".png")),
    width = 2600,
    height = 2400,
    res = 300
  )

  print(volcano_plot)

  invisible(dev.off())

  print(volcano_plot)
}
```

## Heatmaps of top DEGs

```{r top-deg-heatmaps, results='markup', fig.show='hold', fig.width=6.8, fig.height=5.4, out.width='80%'}
row_z_score <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(rep(0, length(x)))
  }

  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

make_top_gene_heatmap <- function(
  vsd,
  results_tbl,
  contrast_id,
  metadata_subset,
  top_n = 25,
  row_font_size = 8,
  column_font_size = 8,
  row_name_width_mm = 85
) {
  top_genes <- results_tbl %>%
    dplyr::filter(contrast == contrast_id) %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(gene_id)

  top_genes <- intersect(top_genes, rownames(SummarizedExperiment::assay(vsd)))

  if (length(top_genes) < 2) {
    warning("Too few genes available for heatmap: ", contrast_id)
    return(NULL)
  }

  heatmap_mat <- SummarizedExperiment::assay(vsd)[top_genes, , drop = FALSE]
  heatmap_mat_z <- t(apply(heatmap_mat, 1, row_z_score))

  symbols <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys = rownames(heatmap_mat_z),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  row_labels <- symbols[match(rownames(heatmap_mat_z), names(symbols))]
  row_labels[is.na(row_labels)] <- rownames(heatmap_mat_z)[is.na(row_labels)]
  rownames(heatmap_mat_z) <- make.unique(row_labels)

  annotation_col <- metadata_subset %>%
    dplyr::select(sample_id, condition) %>%
    dplyr::mutate(condition = factor(condition)) %>%
    as.data.frame()

  rownames(annotation_col) <- annotation_col$sample_id
  annotation_col$sample_id <- NULL

  ha <- ComplexHeatmap::HeatmapAnnotation(
    df = annotation_col,
    which = "column",
    col = list(condition = condition_colors)
  )

  ComplexHeatmap::Heatmap(
    heatmap_mat_z,
    name = "Row Z-score",
    col = zscore_colors,
    top_annotation = ha,
    column_title = paste("Top", top_n, "genes:", contrast_id),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = grid::gpar(fontsize = row_font_size),
    row_names_max_width = grid::unit(row_name_width_mm, "mm"),
    column_names_gp = grid::gpar(fontsize = column_font_size),
    column_names_rot = 45,
    heatmap_legend_param = list(title = "Row Z-score")
  )
}

for (cell in c("AEC", "AM")) {
  dds <- if (cell == "AEC") filtered_dds_AEC else filtered_dds_AM
  vsd <- DESeq2::vst(dds, blind = FALSE)

  meta_cell <- metadata_filtered %>%
    dplyr::filter(cell_type == cell)

  contrast_ids <- unique(filtered_results$contrast[filtered_results$cell_type == cell])

  for (contrast_id in contrast_ids) {
    ht_export <- make_top_gene_heatmap(
      vsd = vsd,
      results_tbl = filtered_results,
      contrast_id = contrast_id,
      metadata_subset = meta_cell,
      top_n = 25,
      row_font_size = 7,
      column_font_size = 7,
      row_name_width_mm = 75
    )

    ht_inline <- make_top_gene_heatmap(
      vsd = vsd,
      results_tbl = filtered_results,
      contrast_id = contrast_id,
      metadata_subset = meta_cell,
      top_n = 25,
      row_font_size = 6.8,
      column_font_size = 6.5,
      row_name_width_mm = 80
    )

    if (!is.null(ht_export) && !is.null(ht_inline)) {
      png(
        file.path(figures_dir, paste0("heatmap_top25_", contrast_id, ".png")),
        width = 2600,
        height = 2200,
        res = 300
      )

      ComplexHeatmap::draw(
        ht_export,
        heatmap_legend_side = "right",
        annotation_legend_side = "right",
        padding = grid::unit(c(8, 8, 10, 16), "mm")
      )

      invisible(dev.off())

      ComplexHeatmap::draw(
        ht_inline,
        heatmap_legend_side = "right",
        annotation_legend_side = "right",
        padding = grid::unit(c(6, 6, 8, 12), "mm")
      )
    }
  }
}
```

# Conclusions

This workflow demonstrates a reproducible RNA-seq analysis and reporting pipeline integrating:

- QC and exploratory diagnostics
- transparent outlier sensitivity analysis
- DESeq2 differential expression workflows
- structured export of annotated DEG outputs
- downstream-ready reporting artifacts for Python-based aggregation and visualization

The analysis emphasizes reproducibility and transparent analytical decision-making
rather than biological overinterpretation.

# Reproducibility information

```{r session-info, echo = TRUE}
sessionInfo()
```
